Using Stochastic Modeling to Predict Long Timescale Transport Behavior of Solutes in an H\textsubscript{II} Phase Lyotropic Liquid Crystal Membrane

### 2019-08-22 Ben_Manuscripts/stochastic_transport/Outline.tex
solved atom naming bug
Responses to some comments

%MRS: better to have a title that described a finding, rather than the
%approach used.  But we may need to wait a bit further until we can decide on a finding.
%MRS: also 'stochastic modeling' is pretty vague.  should be more precise. 
\title{Using Stochastic Modeling to Predict Long Timescale Transport Behavior of Solutes in an H\textsubscript{II} Phase Lyotropic Liquid Crystal Membrane}
%MRS: we should discuss whether Chris Calderon would be on this one.  I would guess no, but I don't know for sure. 
\author{Benjamin J. Coscia \and Michael R. Shirts}

@@

  % MRS: They way to think about it, I think, is that the simplest
  % assumption is the the transport mechanism is simple diffusion;
  % uncorrelated random jumps due to molecular collisions. If you make
  % that assumption, you get MSD linear in time.  So that's IS a
  % stochastic process, but not a very complex one. To make long
  % timescale predictions, we need to increase the detail.  Is that
  % what you are referring to? Not sure if I answered the question . . .

@@

  One can project long timescale solute behavior by modeling transport as a
  stochastic process.
  %MRS: as above, simple diffusion is stochastic, just not very complex.  So we want to add appropriate complexity.
  \begin{itemize}
    %MRS: best talk about these mechanisms in terms of the
    %mathematical models they capture.  But I don't know that it's
    %necessary to review all of them.

@@

  We ran all MD simulations and energy minimizations using GROMACS 2018. We  % needs gromacs citations
  performed all post-simulation trajectory using python scripts which are available
  %MRS: eventually want to say what scripts were used to do what, but that is not at the outline stage!
  online at \\ \texttt{https://github.com/shirtsgroup/LLC\_Membranes}.

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10 \% water by weight. 
  \begin{itemize}
    \item Approximately one third of the water molecules occupy the tail region 
    with the rest near the pore center.
    \item We chose to study the 10 wt \% water system because solutes move 
    significantly faster than in the 5 wt \% system studied previously.
    \item Appropriate stochastic modeling requires that solutes explore 
    as much structural space as possible.  %BJC: better term than structural space?
    %MRS: something simpler like ``sample the accessible mechanisms with representative probability''?
  \end{itemize}

  We chose to study a subset of 4 of the fastest moving solutes from our previous
  work: methanol, acetic acid, urea and ethylene glycol.
  \begin{itemize} 
    \item In addition to exploring membrane structural space the most, these solutes
    have a relatively diverse set of chemical functionality.   
    %MRS: do the solutes also have some diversity of mechanisms (that were observed previously?)
    \item For each solute we created a separate system and to each system we
    added 6 solutes per pore for a total of 24 solutes.
    \item This number of solutes per pore provides a balance of a low 
    degree of interaction between solutes and sufficient amount of data from
    which to generate statistics on the time scales which we simulate.
    \item Further details on the setup and equilibration of these systems can
    be found in our previous work.\cite{coscia_chemically_2019}
  \end{itemize}

@@

  %MRS: this below sort of goes beyond standard use of MSM's though, right? where everything is usually constant-rate expoential processes? Should be explicitly mentioned.
  \noindent We modeled the emission distributions as L\'evy stable distributions.  % cauchy is a case of a Levy stable distribution. The tails are too heavy though.
  \begin{itemize}
    \item For independent and identically distributed (iid) random variables, the 
    generalized central limit theorem guarantees convergence of the associated 
    probability distribution function (PDF) to a L\'evy stable PDF. \cite{klages_anomalous_2008}
    % BJC: ``For sums of independent, iid random variables with proper normalization to the
%    sample size, the generalized central limit theorem guarantees the convergence of
%    the associated PDF to a L\'evy stable PDF even though the variance of these random
%    variables diverges." \cite{klages_anomalous_2008}  % BJC: reword
    \item The characteristic equation describing the Fourier transform of a L\'evy 
    stable PDF is given below:
    \begin{equation}
    p_{\alpha, \beta}(k;\mu,\sigma) =\exp\left[i\mu k - \sigma^{\alpha}|k|^{\alpha}\left(1 - i\beta\frac{k}{|k|}\omega(k, \alpha)\right)\right]
    \end{equation}
    \item Where $\omega$ is defined with something that is complicated to format and can be 
    added later if I keep these formulas in. % https://tex.stackexchange.com/questions/32140/how-to-write-a-function-piecewise-with-bracket-outside
    \item $\alpha$ is the index of stability or L\'evy index, $\beta$ is the skewness 
    parameter, $\mu$ is the shift parameter and $\sigma$ is a scale parameter.
    \item The most familiar case, and 1 of 3 that can be expressed in terms of elementary
    functions, is the Gaussian PDF ($\alpha$ = 2).
    \item The more general family of L\'evy stable distributions allows
    greater flexibility in defining the observed emission distribution PDFs.
    \item In the case that the empirical emission distributions are iid, sequential draws
    from a L\'evy stable PDF defines a L\'evy process. 
    \item We can relax the iid assumption and maintain the flexibility of L\'evy stable 
    probability distributions if we instead consider the dynamics to be governed by a
    non-Gaussian Ornstein-Uhlenbeck (OU) process.\cite{barndorffnielsen_non-gaussian_2001} % BJC: replaces Wiener process with Levy process
    \item OU processes contain temporal dependence but the marginal observation distributions
    are consistent with the underlying stochastic driver.~\cite{taufer_simulation_2009}
    \item We will discuss the implications of each.  
    %MRS: does the process described above include the motion within the state? That is something I feel that the
    %standard MSM doesn't include either, since they are typically interested just in the distribution of states in equilibrium, not the diffusion. 
  \end{itemize}

  L\'evy stable distributions often have heavy tails and an undefined variance 
  which can give rise to arbitrarily long particle displacements.
  \begin{itemize}
  	\item We do not observe hops longer than x nm in our simulations which 
  	emphasizes that these distributions are only an approximation.
    \item We avoid enormous and unrealistic hops problem by truncating the tails
    of the distribution. %BJC: can add some citations about truncated Levy distributions
    \item Since solutes in our system cannot hop further than the length of our 
    simulation unit cell, we truncate the tails of the distribution at this
    magnitude.
    % BJC: Is this actually the right way to do this? I am looking at unwrapped coordinates,
    % so theoretically a particle can move as far as it wants between frames.
    % BJC: The cut-off choice is pretty important and has a significant effect on MSD. 
    % For example, if I truncate at the unit cell length I reproducably get a final MSD of ~ 66 nm^2 (1000 ns stochastic simulation)
    % If I truncate the tails at double the unit cell length, I get a final MSD of ~107 nm^2
    % The uncertainty also increases (since variance diverges for these distributions)
    % So maybe it makes the most sense to truncate based on largest observed hops.
  \end{itemize}

@@

  %MRS: you intend this to be a separate model than the above?  If so, make clearer.
  \subsection{Modeling subdiffusion}\label{method:model_sFBM}

  Solutes in this system exhibit subdiffusive behavior, a type of anomalous diffusion.
  \begin{itemize}
  %MRS: you would probably want to be clear what the mechanistic reasons for for anomalous diffusion: 
  %want to talk about why, given the mechanisms you would consider. 
  	\item During an anomalous diffusion process, the mean squared displacement (MSD)
  	does not grow linearly with time, rather it is of the form:
	\begin{equation} 
	\langle x^2(t) \rangle = K_{\alpha}t^\alpha
	\label{eqn:msd_form}
	\end{equation} 
	where $\alpha$ is the anomalous exponent and $K_\alpha$ is the
	generalized diffusion coefficient.
	\item A value of $\alpha < 1$ indicates a subdiffusive process, while values of
	$\alpha = 1$ and $\alpha > 0$ are characteristic of Brownian and superdiffusive
	motion respectively.
  \end{itemize}

  \noindent We analyzed both the ensemble-averaged and time-averaged MSDs
  of the simulated trajectories.
  \begin{itemize}
    %MRS: describe the physical situations in which one would want to choose each of these ways of calculating the MSD.
	\item The ensemble-averaged MSD measures the displacement of a particle from its initial
	position~\cite{meroz_toolbox_2015} and can be written as
	\begin{equation}
	\langle x^2(t) \rangle = \langle x(t) - x(0) \rangle^2
	\label{eqn:ensemble_msd}
	\end{equation}
%	\item The average MSD calculated in this way recovers the form of
%	Equation~\ref{eqn:msd_form}. % time-averaged should recover it too for FBM 
	\item The time-averaged MSD measures the displacement between all possible time lags
	and can be written as
	\begin{equation}
	\overline{x^2(\tau)} = \dfrac{1}{T - \tau}\int_{0}^{T - \tau} (x(t + \tau) - x(t))^2 dt
	\end{equation}
	where $\tau$ is the time lag and T is the length of the
	trajectory~\cite{meroz_toolbox_2015}. 
  \end{itemize}

  \noindent Three common mathematical models for modeling anomalous subdiffusion
  processes include continuous time random walks (CTRW), fractional Brownian motion
  (FBM) and random walks on fractals (RWF).\cite{meroz_toolbox_2015}
  \begin{itemize}
    \item FBM is common in crowded, viscoelastic environments where each step comes 
    from a Gaussian distribution but is anti-correlated to its previous 
    step.~\cite{mandelbrot_fractional_1968,jeon_fractional_2010,banks_anomalous_2005}
    \item A CTRW is characterized by a distribution of hop lengths and 
    dwell times, where each each step is characterized by independent random draws from 
    each distribution.\cite{montroll_random_1965,morrin_three_2018}
    \item An RWF is imposed by a system's geometry. Systems with tortuous pathways and dead
    ends cause anti-correlated motion.\cite{meroz_toolbox_2015,neusius_subdiffusion_2008}
    \item The processes described above can happen alone or in combination.  	
  \end{itemize}
  %MRS: talk about which of these should apply given the data mechanistically, as well as fit.

@@

  % BJC: need better or more descriptive way of describing stochastic simulations
  % BJC: or could say that whenever we say simulations, we mean stochastic simulations unless it is specified as a molecular dynamics simulation.
  % MRS: could say timeseries simulation?  MD is stochastic as well.
  Simulating solute trajectories (see Section~\ref{method:MSMs}) by drawing independent  
  observations from the emission distributions yields near linear MSD curves
  with final values that are far higher than those observed in our molecular simulations.
  \begin{itemize}
	% BJC: right number of solute trajectories to simulate for comparison? More tightens the error bars
    \item In the case of methanol, the final MSD value is about 25x higher than that observed in our
    previous transport study (see Figure~\ref{fig:msd_MET}).
    \item The linear behavior of the MSD curve is not consistent with subdiffusion.
  \end{itemize}
  %MRS: so, is there a way we should have been able to tell beforrehand that this approach didn't work that we an explain?
  %or can we show we needed to do this to get the results?
  \begin{figure}[!h]
  \centering
  \includegraphics[width=0.5\textwidth]{MET_msd.pdf}
  \caption{}\label{fig:msd_MET}
  \end{figure}

>>>>

%MRS: better to have a title that described a finding, rather than the
%approach used.  But we may need to wait a bit further until we can decide on a finding.
%MRS: also 'stochastic modeling' is pretty vague.  should be more precise. 
\title{Using Stochastic Modeling to Predict Long Timescale Transport Behavior of Solutes in an H\textsubscript{II} Phase Lyotropic Liquid Crystal Membrane}
%MRS: we should discuss whether Chris Calderon would be on this one.  I would guess no, but I don't know for sure. 
%BJC1: Yea maybe Chris as an acknowledgement on this one. He didn't directly contribute anything other than pointing
%me towards Greg Morrin who pointed me towards useful anomalous diffusion literature.
\author{Benjamin J. Coscia \and Michael R. Shirts}

@@

  % MRS: They way to think about it, I think, is that the simplest
  % assumption is the the transport mechanism is simple diffusion;
  % uncorrelated random jumps due to molecular collisions. If you make
  % that assumption, you get MSD linear in time.  So that's IS a
  % stochastic process, but not a very complex one. To make long
  % timescale predictions, we need to increase the detail.  Is that
  % what you are referring to? Not sure if I answered the question . . . 
  % BJC1: Yea, I see what you're saying. Just not sure what else the long
  % timescale will tell us that we can't learn from short ones. Probably 
  % better to talk about this in person.

@@

  We can use the observed mechanisms in order to choose a stochastic model of
  sufficient complexity One can project long timescale solute behavior by modeling transport as a
  stochastic process.
  %MRS: as above, simple diffusion is stochastic, just not very complex.  So we want to add appropriate complexity.
  \begin{itemize}
    %MRS: best talk about these mechanisms in terms of the
    %mathematical models they capture.  But I don't know that it's
    %necessary to review all of them.

@@

  We ran all MD simulations and energy minimizations using GROMACS 2018. We  % needs gromacs citations
  performed all post-simulation trajectory using python scripts which are available
  %MRS: eventually want to say what scripts were used to do what, but that is not at the outline stage!
  %BJC1: I usually put a table like that in the SI. Is that sufficient?
  online at \\ \texttt{https://github.com/shirtsgroup/LLC\_Membranes}.

@@

  We studied transport of solutes in the H\textsubscript{II} phase using an
  atomistic molecular model of four pores in a monoclinic unit cell with 
  10 \% water by weight. 
  \begin{itemize}
    \item Approximately one third of the water molecules occupy the tail region 
    with the rest near the pore center.
    \item We chose to study the 10 wt \% water system because solutes move 
    significantly faster than in the 5 wt \% system studied previously.
    \item Appropriate stochastic modeling requires that solutes sample the 
    accessible mechanisms with representative probability.  %BJC1: incorporated suggested phrasing
  \end{itemize}

  We chose to study a subset of 4 of the fastest moving solutes from our previous
  work: methanol, acetic acid, urea and ethylene glycol.
  \begin{itemize} 
    \item In addition to exploring membrane structural space the most, these solutes
    have a relatively diverse set of chemical functionality.   
    %MRS: do the solutes also have some diversity of mechanisms (that were observed previously?)
    %BJC1: Yes. Need to put following into words
    %BJC1: Ranking of time spent in pore region: AcOH > EG > URE > MeOH  (spans 4th most to 3rd least)
    %BJC1: All have high hop frequency which is consistent with fast motion
    %BJC1: Hbonding: Urea barely hbonds. AcOH and EG hbond a lot. AcOH has highest hbond lifetime of all 20 solutes. MeOH on par with all other simple alcohols
    %BJC1: Sodium Association: Urea coordinates with sodium the most of all 20 solutes. AcOH, EG, MeOH spread out of rest of range. Lifetime similar 
    \item For each solute we created a separate system and to each system we
    added 6 solutes per pore for a total of 24 solutes.
    \item This number of solutes per pore provides a balance of a low 
    degree of interaction between solutes and sufficient amount of data from
    which to generate statistics on the time scales which we simulate.
    \item Further details on the setup and equilibration of these systems can
    be found in our previous work.\cite{coscia_chemically_2019}
  \end{itemize}

@@

  % MRS: this below sort of goes beyond standard use of MSM's though, right? where everything is 
  % usually constant-rate expoential processes? Should be explicitly mentioned.N
  %BJC1: yes
  \noindent We modeled the emission distributions as L\'evy stable distributions.  % cauchy is a case of a Levy stable distribution. The tails are too heavy though.
  \begin{itemize}
    \item For independent and identically distributed (iid) random variables, the 
    generalized central limit theorem guarantees convergence of the associated 
    probability distribution function (PDF) to a L\'evy stable PDF. \cite{klages_anomalous_2008}
    % BJC: ``For sums of independent, iid random variables with proper normalization to the
%    sample size, the generalized central limit theorem guarantees the convergence of
%    the associated PDF to a L\'evy stable PDF even though the variance of these random
%    variables diverges." \cite{klages_anomalous_2008}  % BJC: reword
    \item The characteristic equation describing the Fourier transform of a L\'evy 
    stable PDF is given below:
    \begin{equation}
    p_{\alpha, \beta}(k;\mu,\sigma) =\exp\left[i\mu k - \sigma^{\alpha}|k|^{\alpha}\left(1 - i\beta\frac{k}{|k|}\omega(k, \alpha)\right)\right]
    \end{equation}
    \item Where $\omega$ is defined with something that is complicated to format and can be 
    added later if I keep these formulas in. % https://tex.stackexchange.com/questions/32140/how-to-write-a-function-piecewise-with-bracket-outside
    \item $\alpha$ is the index of stability or L\'evy index, $\beta$ is the skewness 
    parameter, $\mu$ is the shift parameter and $\sigma$ is a scale parameter.
    \item The most familiar case, and 1 of 3 that can be expressed in terms of elementary
    functions, is the Gaussian PDF ($\alpha$ = 2).
    \item The more general family of L\'evy stable distributions allows
    greater flexibility in defining the observed emission distribution PDFs.
    \item In the case that the empirical emission distributions are iid, sequential draws
    from a L\'evy stable PDF defines a L\'evy process. 
    \item We can relax the iid assumption and maintain the flexibility of L\'evy stable 
    probability distributions if we instead consider the dynamics to be governed by a
    non-Gaussian Ornstein-Uhlenbeck (OU) process.\cite{barndorffnielsen_non-gaussian_2001} % BJC: replaces Wiener process with Levy process
    \item OU processes contain temporal dependence but the marginal observation distributions
    are consistent with the underlying stochastic driver.~\cite{taufer_simulation_2009}
    \item We will discuss the implications of each.  
    %MRS: does the process described above include the motion within the state? That is something I feel that the
    %standard MSM doesn't include either, since they are typically interested just in the distribution of states in equilibrium, not the diffusion. 
  \end{itemize}

  L\'evy stable distributions often have heavy tails and an undefined variance 
  which can give rise to arbitrarily long particle displacements.
  \begin{itemize}
  	\item We do not observe hops longer than x nm in our simulations which 
  	emphasizes that these distributions are only an approximation.
    \item We avoid enormous and unrealistic hops problem by truncating the tails
    of the distribution. %BJC: can add some citations about truncated Levy distributions which are used frequently in modeling
    \item Since solutes in our system cannot hop further than the length of our 
    simulation unit cell, we truncate the tails of the distribution at this
    magnitude.
    % BJC: Is this actually the right way to do this? I am looking at unwrapped coordinates,
    % so theoretically a particle can move as far as it wants between frames.
    % BJC: The cut-off choice is pretty important and has a significant effect on MSD. 
    % For example, if I truncate at the unit cell length I reproducably get a final MSD of ~ 66 nm^2 (1000 ns stochastic simulation)
    % If I truncate the tails at double the unit cell length, I get a final MSD of ~107 nm^2
    % The uncertainty also increases (since variance diverges for these distributions)
    % So maybe it makes the most sense to truncate based on largest observed hops.
  \end{itemize}

@@

  %MRS: you intend this to be a separate model than the above?  If so, make clearer.
  %BJC1: renamed, is that what you meant?
  \subsection{The Anomalous Diffusion Model}\label{method:model_sFBM}

  Solutes in this system exhibit subdiffusive behavior, a type of anomalous diffusion.
  \begin{itemize}
  %MRS: you would probably want to be clear what the mechanistic reasons for for anomalous diffusion: 
  %want to talk about why, given the mechanisms you would consider. 
  %BJC1: could go through flow chart
  	\item During an anomalous diffusion process, the mean squared displacement (MSD)
  	does not grow linearly with time, rather it is of the form:
	\begin{equation} 
	\langle x^2(t) \rangle = K_{\alpha}t^\alpha
	\label{eqn:msd_form}
	\end{equation} 
	where $\alpha$ is the anomalous exponent and $K_\alpha$ is the
	generalized diffusion coefficient.
	\item A value of $\alpha < 1$ indicates a subdiffusive process, while values of
	$\alpha = 1$ and $\alpha > 0$ are characteristic of Brownian and superdiffusive
	motion respectively.
  \end{itemize}

  \noindent We analyzed both the ensemble-averaged and time-averaged MSDs
  of the simulated trajectories.
  \begin{itemize}
    %MRS: describe the physical situations in which one would want to choose each of these ways of calculating the MSD.
    %BJC1: For determining ergodicity. Ensemble for determining anomalous exponent. Not sure I'll use ensemble MSD
	\item The ensemble-averaged MSD measures the displacement of a particle from its initial
	position~\cite{meroz_toolbox_2015} and can be written as
	\begin{equation}
	\langle x^2(t) \rangle = \langle x(t) - x(0) \rangle^2
	\label{eqn:ensemble_msd}
	\end{equation}
%	\item The average MSD calculated in this way recovers the form of
%	Equation~\ref{eqn:msd_form}. % time-averaged should recover it too for FBM 
	\item The time-averaged MSD measures the displacement between all possible time lags
	and can be written as
	\begin{equation}
	\overline{x^2(\tau)} = \dfrac{1}{T - \tau}\int_{0}^{T - \tau} (x(t + \tau) - x(t))^2 dt
	\end{equation}
	where $\tau$ is the time lag and T is the length of the
	trajectory~\cite{meroz_toolbox_2015}. 
  \end{itemize}

  \noindent Three common mathematical models for modeling anomalous subdiffusion
  processes include continuous time random walks (CTRW), fractional Brownian motion
  (FBM) and random walks on fractals (RWF).\cite{meroz_toolbox_2015}
  \begin{itemize}
    \item FBM is common in crowded, viscoelastic environments where each step comes 
    from a Gaussian distribution but is anti-correlated to its previous 
    step.~\cite{mandelbrot_fractional_1968,jeon_fractional_2010,banks_anomalous_2005}
    \item A CTRW is characterized by a distribution of hop lengths and 
    dwell times, where each each step is characterized by independent random draws from 
    each distribution.\cite{montroll_random_1965,morrin_three_2018}
    \item An RWF is imposed by a system's geometry. Systems with tortuous pathways and dead
    ends cause anti-correlated motion.\cite{meroz_toolbox_2015,neusius_subdiffusion_2008}
    \item The processes described above can happen alone or in combination.  	
  \end{itemize}
  %MRS: talk about which of these should apply given the data mechanistically, as well as fit.
  %BJC1: I believe that's what I try to do in the paragraph below. I can be a little more explicit

@@

  % BJC: need better or more descriptive way of describing stochastic simulations
  % BJC: or could say that whenever we say simulations, we mean stochastic simulations unless it is specified as a molecular dynamics simulation.
  % MRS: could say timeseries simulation?  MD is stochastic as well.
  Simulating solute trajectories (see Section~\ref{method:MSMs}) by drawing independent  
  observations from the emission distributions yields near linear MSD curves
  with final values that are far higher than those observed in our molecular simulations.
  \begin{itemize}
	% BJC: right number of solute trajectories to simulate for comparison? More tightens the error bars
    \item In the case of methanol, the final MSD value is about 25x higher than that observed in our
    previous transport study (see Figure~\ref{fig:msd_MET}).
    \item The linear behavior of the MSD curve is not consistent with subdiffusion.
  \end{itemize}
  %MRS: so, is there a way we should have been able to tell beforrehand that this approach didn't work that we an explain?
  %or can we show we needed to do this to get the results?
  %BJC1: could just show autocorrelation function of all increments, which has negative dip at k=1 
  \begin{figure}[!h]
  \centering
  \includegraphics[width=0.5\textwidth]{MET_msd.pdf}
  \caption{}\label{fig:msd_MET}
  \end{figure}
